---
title: "Explore Model Specifications"
format: html
---

```{r}
library(tidyverse)
library(here)
library(sandwich)
library(lmtest)
```

```{r}
# load exploration dataset
source_file <- here("data", "processed", "childcare_explore.csv")

df <- read_csv(source_file)
```

```{r}
glimpse(df)
```

```{r}
df <- df |> 
  mutate(
    county_fips_code = as.factor(county_fips_code),
    county_name = as.factor(county_name),
    state_name = as.factor(state_name),
    state_abbreviation = as.factor(state_abbreviation)
  )
```


# Operationalization of childcare price

```{r}
df |> 
  select(county_fips_code, mcinfant, mctoddler, mcpreschool) |> 
  summary()
```

```{r}
# create boxplot
df |> 
  select(county_fips_code,mcinfant, mctoddler, mcpreschool) |> 
  pivot_longer(
    cols = starts_with("mc"),
    names_to = "age_groups",
    values_to = "childcare_med_price"
  ) |> 
  mutate(
    age_groups = factor(age_groups, levels = c("mcinfant", "mctoddler", "mcpreschool"))
  ) |> 
  ggplot(aes(x = age_groups, y = childcare_med_price, colour = age_groups)) +
  geom_boxplot() +
  theme(
    legend.position = "none"
  )
```


### Infant and Toddler childcare price

```{r}
t.test(df$mcinfant, df$mctoddler, paired = TRUE)
```


### Infant and Preschool childcare price

```{r}
t.test(df$mcinfant, df$mcpreschool, paired = TRUE)
```

### Toddler and Preschool childcare price

```{r}
t.test(df$mctoddler, df$mcpreschool, paired = TRUE)
```

# Plots: individual var

## Childare price

```{r}
hist(df$mcinfant, breaks = 20)
```
```{r}
hist(df$mctoddler, breaks = 20)
```
```{r}
hist(df$mcpreschool, breaks = 20)
```

## Market size

```{r}
hist(df$market_size, breaks = 100 )
```
## Household Income

```{r}
hist(df$mhi, breaks = 75)
```
```{r}
hist(df$mhi_2022, breaks = 75)
```


# Scatterplot: Shape of relationships

## Infant childcare price and market size

```{r}
# level-level
df |> 
  ggplot(aes(x = market_size, y = mcinfant)) +
  geom_point(color = "blue", alpha = 0.6)
```
```{r}
# x: level, y: log (natural)
df |> 
  ggplot(aes(x = market_size, y = log(mcinfant))) +
  geom_point(color = "blue", alpha = 0.6)
```
```{r}
# x: log (natural), y: level
df |> 
  ggplot(aes(x = log(market_size), y = mcinfant)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method="lm", color = "red")
```
```{r}
# x: log (natural), y: log(natural)
df |> 
  ggplot(aes(x = log(market_size), y = log(mcinfant))) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method="lm", color = "red")
```
## Infant childcare price and household income

```{r}
# level-level
df |> 
  ggplot(aes(x = mhi_2022, y = mcinfant)) +
  geom_point(color = "darkgreen", alpha = 0.6)
```
```{r}
# x: level, y: log (natural)
df |> 
  ggplot(aes(x = mhi_2022, y = log(mcinfant))) +
  geom_point(color = "darkgreen", alpha = 0.6) +
  geom_smooth(method="lm", color = "red")
```
```{r}
# x: log (natural), y: level
df |> 
  ggplot(aes(x = log(mhi_2022), y = mcinfant)) +
  geom_point(color = "darkgreen", alpha = 0.6)
```
```{r}
# x: log (natural), y: log(natural)
df |> 
  ggplot(aes(x = log(mhi_2022), y = log(mcinfant))) +
  geom_point(color = "darkgreen", alpha = 0.6) +
  geom_smooth(method="lm", color = "red")
```


# Models

## Model 1: Base model

### Model 1a: Infant

```{r}
model_1a <- lm(log(mcinfant) ~ log(market_size), data = df)
coeftest(model_1a, vcov = vcovHC(model_1a))
```

```{r}
plot(model_1a)
```


### Model 1b: Toddler

```{r}
model_1b <- lm(log(mctoddler) ~ log(market_size), data = df)
coeftest(model_1b, vcov = vcovHC(model_1b))
```

```{r}
plot(model_1b)
```

### Model 1c: Preschool

```{r}
model_1c <- lm(log(mcpreschool) ~ log(market_size), data = df)
coeftest(model_1c, vcov = vcovHC(model_1c))
```

```{r}
plot(model_1c)
```


## Model 2: with level-household income

### Model 2a: Infant

```{r}
model_2a <- lm(log(mcinfant) ~ log(market_size) + mhi_2022, data = df)
coeftest(model_2a, vcov = vcovHC(model_2a))
```

```{r}
plot(model_2a)
```

### Model 2b: Toddler

```{r}
model_2b <- lm(log(mctoddler) ~ log(market_size) + mhi_2022, data = df)
coeftest(model_2b, vcov = vcovHC(model_2b))
```

```{r}
plot(model_2b)
```

### Model 2c: Preschool

```{r}
model_2c <- lm(log(mcpreschool) ~ log(market_size) + mhi_2022, data = df)
coeftest(model_2c, vcov = vcovHC(model_2c))
```

```{r}
plot(model_2c)
```



## Model 3: with log-household income

### Model 3a: Infant

```{r}
model_3a <- lm(log(mcinfant) ~ log(market_size) + log(mhi_2022), data = df)
coeftest(model_3a, vcov = vcovHC(model_3a))
```

```{r}
plot(model_3a)
```


### Model 3b: Toddler

```{r}
model_3b <- lm(log(mctoddler) ~ log(market_size) + log(mhi_2022), data = df)
coeftest(model_3b, vcov = vcovHC(model_3b))
```

```{r}
plot(model_3b)
```

### Model 3c: Preschool

```{r}
model_3c <- lm(log(mcpreschool) ~ log(market_size) + log(mhi_2022), data = df)
coeftest(model_3c, vcov = vcovHC(model_3c))
```

```{r}
plot(model_3c)
```


